---
title: "Class 19: Genome informatics"
author: "Andrew Kapinos"
date: "12/1/2021"
output: github_document
---
>Q13: Read this file into R and determine the sample size for each genotype and their corresponding median expression levels for each of these genotypes.

```{r message=FALSE}
data <- read.table("rs8067378_ENSG00000172057.6.txt")
table(data$geno)
print(paste("The sample size for genotype A/A is", table(data$geno)["A/A"]))
print(paste("The sample size for genotype A/G is", table(data$geno)["A/G"]))
print(paste("The sample size for genotype G/G is", table(data$geno)["G/G"]))

library(dplyr)
summary(data[data$geno=="A/A",]$exp)
print(paste("The median expression value for genotype A/A is",
            round(summary(data[data$geno=="A/A",]$exp)["Median"],3)))
summary(data[data$geno=="A/G",]$exp)
print(paste("The median expression value for genotype A/G is",
            round(summary(data[data$geno=="A/G",]$exp)["Median"],3)))
summary(data[data$geno=="G/G",]$exp)
print(paste("The median expression value for genotype G/G is", 
            round(summary(data[data$geno=="G/G",]$exp)["Median"],3)))
```

>Q14: Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3?

```{r message=FALSE}
library(ggplot2)
ggplot(data,aes(geno,exp)) +
  geom_boxplot(notch=TRUE,aes(fill=geno),outlier.shape=NA) +
  geom_point(alpha=0.4,position=position_jitter(0.2)) +
  labs(x="Genotype", y="Expression") +
  theme(legend.position="none")
t.test(data[data$geno=="A/A",]$exp,data[data$geno=="G/G",]$exp)
```

On average, the A/A genotype results in **relatively higher** expression levels when compared to the G/G genotype. Given the results of the boxplot, it would appear as though the SNP does in fact affect expression of ORMDL3.

A Student's t-test confirms that the difference in means is statistically significant, with a p-value of 2.2e-16.
