---
title: "Improving analysis code by writing functions"
author: "Andrew Kapinos"
date: "10/23/2021"
output: github_document
---

# Can you improve this analysis code?
library(bio3d)
s1 <- read.pdb("4AKE") # kinase with drug
s2 <- read.pdb("1AKE") # kinase no drug
s3 <- read.pdb("1E4Y") # kinase with drug

s1.chainA <- trim.pdb(s1, chain="A", elety="CA")
s2.chainA <- trim.pdb(s2, chain="A", elety="CA")
s3.chainA <- trim.pdb(s3, chain="A", elety="CA")

s1.b <- s1.chainA$atom$b
s2.b <- s2.chainA$atom$b
s3.b <- s3.chainA$atom$b

plotb3(s1.b, sse=s1.chainA, typ="l", ylab="Bfactor")
plotb3(s2.b, sse=s2.chainA, typ="l", ylab="Bfactor")
plotb3(s3.b, sse=s3.chainA, typ="l", ylab="Bfactor")

> Q6. How would you generalize the original code above to work with any set of input protein structures?

We need to break down the original code into its various steps.

Step 1: Load bio3d packages (not neccessary to include in function).
Step 2: Access online PDB file and assign to an object.
Step 3: Trim PDB file to only include chain A.
Step 4: Describe b for atoms in chain A.
Step 5: Plot data.

Starting at step 2, we can simplify the code given for accessing the PDB files and assigning them to an object. At each step, we'll run the original code and our new version to ensure that the same output is generated.

```{r}
# Original code
library(bio3d)
s1 <- read.pdb("4AKE") # kinase with drug
s1

# Simplified version
library(bio3d)
x <- "4AKE"
x <- read.pdb(x)
x
```
Moving onto step 3, let's simplify the code used to trim the PDB file to only chain A.

```{r}
# Original code
s1.chainA <- trim.pdb(s1, chain="A", elety="CA")
s1.chainA

# Simplified version
chainA <- trim.pdb(x, chain="A", elety="CA")
chainA
```

Moving onto step 4, let's simplify the code used to describe *b* for atoms in chain A.

```{r}

# Original code
s1.b <- s1.chainA$atom$b
s1.b

# Simplified version
b <- chainA$atom$b
b
```

Finally, in step 5, we simplify the code used to plot the data.

```{r}
# Original code
plotb3(s1.b, sse=s1.chainA, typ="l", ylab="Bfactor")

# Simplified version
plotb3(b, sse=chainA, typ="l", ylab="Bfactor")
```

Let's put it all together!

```{r}
x <- "4AKE"
library(bio3d)
x <- read.pdb(x)
chainA <- trim.pdb(x, chain="A", elety="CA")
b <- chainA$atom$b
plotb3(b, sse=chainA, typ="l", ylab="Bfactor")
```

Now, let's write our code into a function.

```{r}
prot_drug_interaction <- function(x) {
  library(bio3d)
  x <- read.pdb(x)
  chainA <- trim.pdb(x, chain="A", elety="CA")
  b <- chainA$atom$b
  plotb3(b, sse=chainA, typ="l", ylab="Bfactor")
}
```

And let's verify that it works for of our sample input.
```{r}
prot_drug_interaction("4AKE")
```

Let's create a data frame that contains the PDB codes provided as example inputs above. Then, we can use *apply()* to run the function on each of the original input PDB codes, by checking each row in the data frame for the input, x (ie. the PDB code).

```{r}
PDB_codes <- data.frame(c("4AKE","1AKE","1E4Y"))
apply(PDB_codes,1,prot_drug_interaction)
```

